Tutorial: Fitting log-Gaussian mixed-effects mixture models for keystroke data

### 2024-05-02 tutorial/tutorial.Rmd
update tutorial final for now

```{r, include=FALSE}
library(rmdformats)
library(knitr)
library(tidyverse)
library(rstan)
library(brms)
```

@@

@@

@@

The priors default priors and the priors we are going to set are now no longer on the msecs scale but on the log scale.

@@

For example, a mean of 5 msecs on the log scale are `exp(5.85)` $\approx$ `r round(exp(5.85),0)` msecs on a linear scale. This prior with varying degrees of freedom is visualised in Figure \@ref(fig:logprior).

@@

## Model comparison

To evaluate the predictive performance of the models we can use leave-one-out cross-validation. The out-of-sample predictive performance was determined via Pareto smoothed importance-sampling [@vehtari2015pareto; @vehtari2017practical]. Cross-validation, in contrast to more conventional model-comparison metrics such as $R^2$, prevent overfitting by penalising models with more parameters [see @mcelreath2016statistical; @lambert2018student].

Cross validation generally is measuring the performance of a model, meaning its ability to capture new data. Overfitting happens when a model is capturing nuances of a sample, rather than patterns one would expect to repeat in new samples. The leave-one-out information criterion (LOO-IC) is determined as follows: train a model on $N-1$ observations, predict the remaining data points from the training model, repeat process $N$ times to predict every observation from a model of th remaining data. Adding up the prediction results gives an estimate of the expected log predictive density (`elpd_loo`; $\widehat{elpd}$), i.e. an approximation of the results that would be expected for new data.

The function `loo` can be used to access the expected log predictive density which uses probability calculations for approximate LOO-IC vio Pareto smoothed importance sampling [@vehtari2015pareto; @vehtari2017practical].

Approximation involved in `loo` uses the log posterior predictive densities: how likely is each data point given the distribution parameter estimates? This is shown for each data point in Figure \@ref(fig:loglik). Extremely small values indicate that an observation is very unlike for a given set of parameter values.

```{r loglik, fig.cap="Log-posterior densities of each data point"}
log_lik(fit_gaus) %>% 
  as_tibble() %>% 
  mutate(sample = 1:n()) %>%
  pivot_longer(-sample, 
               names_to = "obs", 
               values_to = "loglik") %>%
  summarise(M = mean(loglik), var = var(loglik), .by = obs) %>%
  filter(M > -10) %>% # there are some terrible ones for this model 
  mutate(obs = as.numeric(str_remove(obs, "V"))) %>%
  ggplot(aes(x = obs, y = M, ymin = M - var, ymax = M + var)) +
  geom_pointrange(fatten = .5) +
  labs(x = "Observations", 
       y = "log-posterior predictive density (with variance)")
```

`elpd_loo` is the sum of the means and returns the expected log predictive density. `p_loo` is the sum of the variances and indicates the flexibility of the model fit (also used as the effective number of parameters). `looic`, or LOO-IC, corresponds to $-2 \cdot ($`elpd_loo`$-$`p_loo`$)$ (for deviance scale). This can be manually estimated

```{r}
log_lik(fit_gaus) %>% 
  as_tibble() %>% 
  mutate(sample = 1:n()) %>%
  pivot_longer(-sample, 
               names_to = "obs", 
               values_to = "loglik") %>%
  summarise(mean_loglik = mean(exp(loglik)),
            var_loglik = var(loglik),
            .by = obs) %>% 
  summarise(elpd_loo = sum(log(mean_loglik)), # sum of mean log lik
            p_loo = sum(var_loglik), # effective number of parameters
            looic = -2 * (elpd_loo - p_loo)) %>% # Bayesian deviance
  pivot_longer(everything(), names_to = "Estimate") %>% 
  mutate(across(value, round, 1))
```

or directly calculated using `loo`. The Pareto k diagnostic indicates that there are some observations that are very unusual under the single distribution Gaussian model.

@@

@@

```{r}
loo_fit <- loo(fit_gaus, fit_lgaus)
loo_fit$diffs %>% 
  as.data.frame() %>% 
  round(1)
```
The model comparison shows that the model `fit_lgaus` (the model with the log-Gaussian likelihood) is approximately `r round(abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2]),0)` times better than `fit_gaus` (try `abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2])`).

Aside, `elpd_loo` and `looic` have rarely got a direct interpretation. Important is the difference between models. The size of the values is increasing as a function of the number of observations. Generally `elpd_loo` and `looic` measures how many possible futures the model can rule out.

In contrast, `loo_R2` provides a Bayesian $R^2$ that measures the variability in the data that the model can captures. This quantity is not ideal for model comparisons, as normal $R^2$, because it is subject to overfitting: models with more parameters tend to have a larger $R^2$. A larger $R^2$ means that a model captures the data well but might make predictions that do not generalise out-of-sample [see chapter 11 in @gelman2020regression; chapter 6 in @mcelreath2016statistical]. The is an example for Bayesian $R^2$ for the single-distribution log-Gaussian model.

@@

The model returns population estimates for the average typing speed of fluent keystroke transitions ($\beta$), the slowdown for hesitant transitions ($\delta$) and the frequency of hesitant transitions ($\theta$) for each level of a categorical predictor after taking into account two sources of random error: (1) each participant has an individual fluent typing speed that differs across levels of the categorical predictor; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.

$$
\begin{align}
\text{iki}_{i} \sim & \theta_\text{location[i]} \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i]}) \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
		&\text{where}\\
		&\delta \sim \text{N}(0,1)\\
		&\text{constraint: } \delta > 0
\end{align}
$$

The parameters of this equation are visualised in Figure \@ref(fig:mixturemodel). The illustrate shows the dissociation of two underlying processes that the model is looking for in the data. There are two distributions of which the left one in grey is the distribution of fluent keystroke intervals; the right distribution in green is wider and captures generally longer keystroke intervals (with some overlap between fluent and disfluent transitions).

```{r mixturemodel, echo = F, fig.cap="What's a mixture model for keystroke intervals? The parameters of the mixture model explained visually. Each of the three parameters is determined for each transition location."}

@@

ggplot(data_mm, aes(x)) +
  scale_x_continuous(labels = exp, breaks = log(c(0, 10, 100, 1000, 10000)), limits = log(c(30, 50000))) +
  scale_y_continuous(limits = c(0, .7)) +
  stat_function(geom = "line", fun = plot_mix_comps, colour = "#000000",
                args = list(alpha[1], sigma[1], lam = 1 - theta[1])) +
  stat_function(geom = "line", fun = plot_mix_comps, colour = "#009E73",
                args = list(alpha2[1], sigmap[1], lam = theta[1])) +
  stat_function(geom = "area", fun = plot_mix_comps, fill = "#000000",
                show.legend = F, alpha = .25, 
                args = list(alpha[1], sigma[1], lam = 1 - theta[1])) +
  stat_function(geom = "area", fun = plot_mix_comps, fill = "#009E73",
                show.legend = F, alpha = .25, 
                args = list(alpha2[1], sigmap[1], lam = theta[1])) +
  annotate("label", x=4,y=.6, label="beta*':'~'Quick typing'", size = 4, parse = T) +
  annotate("label", x=7.25,y=.35, label="atop(theta*':'~'Hesitation probability', '\"Pausing\"')", parse = T, size = 4) +
  annotate("label", x=9,y=.15, parse = T, label="delta*':'~'\"Slowdown\"'", size = 4) +
  geom_segment(x = alpha, y = 0.05, xend = alpha2, yend = 0.05, 
               colour = "blue", alpha = .5,
               arrow = arrow(length = unit(0.2, "cm"), 
                             ends = "both", 
                             type = "closed")) +
  geom_segment(x = alpha, y = 0, xend = alpha, yend = Inf, colour = "darkred", alpha = .5) +
  geom_segment(x = alpha2, y = 0, xend = alpha2, yend = Inf, colour = "darkred", alpha = .5) +
  geom_curve(x = 4, y = .57, xend = alpha - .75, yend = 0.1, 
  arrow = arrow(length = unit(0.03, "npc"))) +
  geom_curve(x = 7.5, y = .3, xend = alpha2 + .85, yend = 0.1,
             curvature=-.3,
  arrow = arrow(length = unit(0.03, "npc"))) +
  geom_curve(x = 9, y = .125, xend = 5.75, yend = 0.0255,
             curvature=-.3,
   arrow = arrow(length = unit(0.03, "npc"))) +
  labs(x = "iki [in msecs]", 
       y = "Density",
       caption = "see Roeser et al. (2021)")

@@

The average typing speed of fluent keystroke transitions is captured by the $\beta$ parameter. The distance between fluent transitions and disfluent transitions is captured by $\delta$ (i.e. the distance between the grey and the green distribution). The hesitation probability or the "pausing frequency" is the relative height of the grey and the green distribution and captured by $\theta$. Most keystrokes are, typically, fluent transitions and hence in the grey distribution. The frequency of the minority of disfluent transitions is captured by the height of the green distribution. As this model assumes a finite mixture of two distributions, the probability of fluent transitions is just $1-\theta$; in other words, a pausing probability of $\theta=.4$ means that the probability of fluent transitions was $.6$. If the pausing probability decreases to $.3$ the proportion of fluent transitions had increased to $.7$.

## Model fit

The data must be transformed into a list to feed them into Stan. The information required from the data is determined in the Stan code and has to do with what we need to estimate the model parameters and how the model is implemented. The exert from the Stan code below shows which information is expected as input (e.g. int is integer), how they are named, and what the smallest (i.e. lower) and largest possible value (i.e. upper) are. The full Stan code can be found at the end of this tutorial.

@@

Depending on the data we might have different numbers of participants and conditions. The Stan code is using indices to be able to fit varying numbers for each. In the code below we use `factor` in combination with `as.numeric` to ensure that there are no empty indices. We create vectors with participant identifiers `ppts` (numbers $1$ through $I$ where $I$ is the number of participants) and a numeric identifier `location` for each transition locations (levels: before-sentence = 1, before-word = 2, within-word = 3).  The returned values are indices for the parameters in the model: For example, for the location identifier, `beta[1]` is the population estimate for non-hesitant typing intervals at before-sentence transition locations, `beta[2]` is same for before-word transition locations; `theta[1]` and `theta[2]` are the disfluency probability on the logit scale for before-sentence and before-word transition locations, respectively.

@@

## Load model

The Stan code can be downloaded from [GitHub](https://github.com/jensroes/prowrite-mixture-models/blob/main/tutorial/stan/molg.stan) using `readLines` and `writeLines`. The code will be saved as "molg.stan" (mixture of log-Gaussians) in a directory called "stan" which needs to be available in your current working directory.

```{r eval = F}
# Create an empty directory called "stan"
dir.create("stan")
# Download Stan code
file <- readLines("https://raw.githubusercontent.com/jensroes/prowrite-mixture-models/main/tutorial/stan/molg.stan")
# Save Stan code to current working directory
writeLines(file, "stan/molg.stan")
```

The Stan model can be loaded and translated into an S4 class object using the `stan_model` function and assign it to the object `molg`. The "molg.stan" file needs to be available in the directory "stan" located inside the current working directory.

@@

This code fits the posterior of a categorical predictor with any number of levels. In other words, the posterior that model returns can be used to calculate simple effects, main effects and interactions of a factorial design. We demonstrate below how to calculate a simple effect. The same logic can be used to calculate main effects and interactions. Also, participant estimates are calculated for both mixture components and the mixing proportions by condition. For including more predictors, or removing, e.g., random error terms, or adjusting priors, the user will have to work directly in the Stan code.

This Stan code used is largely based on @sorensen2016bayesian and @vasishth2017. @sorensen2016bayesian presents a detailed tutorial on how to fit Stan models [see also @lambert2018student].

The model is fitting a number of parameters that are not relevant for our inference. To reduce the size of the R object that contains the posterior we can select the parameters of interest and disregard all parameters that are not relevant for our inference.

@@

The `sampling` function applies the model to the data using the information specified above (iterations, warmup, chains, cores, parameters `pars` to be saved). `save_warmup` is set to `FALSE` to discard the warmup samples (which are not used for inference anyway) to reduce the size of the posterior. To allow reproducibility (and because Bayesian models involves random number generation) we set the seed. The seed can be any number but using the same number ensures the use of the same random number. Lastly, the control argument was specified with higher values for `adapt_delta` and `max_treedepth`: using higher values here mean the model runs slower but supports more careful parameter estimation. The model is assigned to the variable `m`.

@@

@@

Running this model will take a while to complete sampling depending on your hardware specifications. The time it took my machine to complete this job can me viewed using `get_elapsed_time`. Therefore it is worth to not use all cores for this to run or to use a dedicated high performance machine.

```{r}
get_elapsed_time(fit_molg)/60 # in mins
```

Inferred posterior estimates shown as average and 95% probability interval can be extracted using the `print` function for each transition location and the three parameter values of interest.

@@

## Model comparsion

Similar to above we will compare the three models -- the mixture of log-Gaussians, and the uni-model Gaussian and log-Gaussian models -- using leave-one-out cross-validation. Because the mixture model was fitted directly in Stan and the uni-model models were fitted in `brms`, we need to extract the fit statistics before comparing the models using functions provided by the package `loo`.

@@

As described above we used leave-one-out cross-validation to compare the mixture of log-Gaussians model to each the single distribution Gaussian and log-Gaussian model. Their difference is shown in `elpd_diff` ($\Delta\widehat{elpd}$) with standard error shown in `se_diff`. Models are ordered from the model with the highest predictive performance to the lowest (i.e. `elpd_loo`).

@@

To extract the model comparisons and fit statistics:

@@

The model comparison shows that the model `fit_molg` (the mixture of log-Gaussian ) is approximately `r round(abs(loo_comp[2,1] / loo_comp[2,2]),0)` times better than `fit_lgaus` (try `abs(loo_comp[2,1] / loo_comp[2,2])`).

Model convergence can be established in two practically simple techniques. First, we can inspect the chains in trace plots which show the parameter estimate across iterations (here after warmup) for each chain. We will look at the population-level parameters (names are determined in the Stan code) and assign those to `pars`.

To create trace plots, we can apply the `stan_trace` function to the model `m` and extract the MCMC chains for the parameters in `pars`. The `alpha` argument makes the colours slightly more transparent. If the chains overlap and look like "fat hairy caterpillars", chains have converged on the same target distribution.

Second, we can calculate the $\hat{R}$ statistic. Successful convergence is reflected in $\hat{R}$ values smaller than 1.05 [@gelman1992]. $\hat{R}$ is similar to the F statistic: it tells us how much bigger the variability between chains is compared to the variability within chains. A value of $\approx 1$ indicates that the variability is essentially identical signifying convergence. For this we can use the `rhat` function applied to the model and the population parameters.

Convergence problems can have many reasons and therefore many solutions: running longer chains (in particular longer warmups), increasing the maximum treedepth and the average acceptance probability (`adapt_delta`), specifying starting values, adjusting priors, constraining parameters, or changing the parametrisation of the model. Severe convergence problems are indicative of a misspecification of the model.

The model used the posterior parameter values to simulate hypothetical data sets. This happened for every iteration for every chains. In other words there is a total of `r scales::comma(total_samples)` simulated data sets (and samples per parameter) available to us because

Using the `as.matrix` function we extract a matrix of `y_tilde`, the simulated data. This returns a matrix with the size `total_samples` $\times$ `nrow(data)` (so `r scales::comma(total_samples)` $\times$ `r scales::comma(nrow(data))`). We use the `sample` function to draw `N` randomly sampled hypothetical data sets. The `ppc_dens_oberlay` function from the `bayesplot` package is then mapping the observed data ($y$; thick blue line) to the simulated data ($y_{rep}$; thin lightblue lines).

# Posterior probability distributions

At the core of Bayesian inference is the posterior probability distribution. For each model parameter we have `r scales::comma(total_samples)` posterior samples that form a posterior probability distribution. These samples are the results of the number of iterations after warmup for all chains.

@@

This calculation and visualisation allows researchers to identify participants that do or do not differ in terms of hesitation frequency, in our example depending on transition locations.

>>>>

```{r, include=FALSE}
library(rmdformats)
library(knitr)
library(patchwork)
library(tidyverse)
library(rstan)
library(brms)
```

@@

The choice of priors here is weakly informative. The intention is to use knowledge that we generally have about how long key intervals usually are to point the model towards the right area in the parameter space by assigning low probabilities to unlikely or even implausible values. Even if we know very little about the duration of keystroke interval, we do know that they shouldn't be longer than a day, or event an hour, or several minutes, because such long intervals are unlikely to reflect active writing.

@@

Before you save the posterior, create a new directory called "stan" where we will save all Stan outputs in this tutorial.

```{r eval = F}
# Create an empty directory called "stan"
dir.create("stan")
```

@@

The default priors and the priors that we need to specity are now no longer on the msecs scale but on the log scale.

@@

For example, a mean of 5 on the log scale are `exp(5.85)` $\approx$ `r round(exp(5.85),0)` msecs on a linear scale. This prior with varying degrees of freedom is visualised in Figure \@ref(fig:logprior).

@@

These can be converted into msecs using the exponential function `exp`.

```{r}
fixef(fit_lgaus)[,-2] %>% 
  exp() %>% round(0)
```

The posterior estimates obtained from the log-Gaussian model are radically different (larger) from the posterior estimates of the Gaussian model which is likely due. This difference is likely due to the missing lower bound at zero in the Gaussian model as we as its inability to give less emphasis to values in the right tail.

```{r}
fixef(fit_gaus)[,-2] %>% 
  round(0)
```

# Model comparison I

To evaluate the out-of-sample predictive performance of the models we can use leave-one-out cross-validation via Pareto smoothed importance-sampling [@vehtari2015pareto; @vehtari2017practical]. Cross-validation is a alternative to more conventional model-comparison metrics such as $R^2$ as it penalises models with more parameters to prevent overfitting [see @mcelreath2016statistical; @lambert2018student]. Overfitting happens when a model is capturing nuances of a sample, rather than patterns one would expect to repeat in new samples.

In contrast, cross-validation is measuring the performance of a model, meaning its ability to generalise to new data. The leave-one-out information criterion (LOO-IC) is determined as follows: train a model on $N-1$ observations, predict the remaining data points from the training model, repeat process $N$ times to predict every observation from a model of th remaining data. Adding up the prediction results gives an estimate of the expected log-predictive density (`elpd_loo`; $\widehat{elpd}$), i.e. an approximation of the results that would be expected for new data.

The function `loo` can be used to access the expected log-predictive density which uses probability calculations to approximate LOO-IC via Pareto smoothed importance sampling [@vehtari2015pareto; @vehtari2017practical]. LOO can be approximated with the log posterior predictive densities as a measure of how likely a data point is, given the parameter estimates. This is shown for each data point in Figure \@ref(fig:loglik). Extremely small values indicate that an observation is very unlike under the model's parameter values and we are unlike to observe them in a new sample.

```{r loglik, fig.cap="Log-posterior densities of each data point (errorbars show variance) of the Gaussian model."}
log_lik(fit_gaus) %>% 
  as_tibble() %>% 
  pivot_longer(cols = everything(),
               names_to = "obs", 
               values_to = "loglik") %>%
  summarise(mean = mean(loglik), var = var(loglik), .by = obs) %>%
  filter(mean > -10) %>% # there are some terrible ones for this model 
  mutate(obs = as.numeric(str_remove(obs, "V"))) %>%
  ggplot(aes(x = obs, y = mean, ymin = mean - var, ymax = mean + var)) +
  geom_pointrange(fatten = .5) +
  labs(x = "observations", 
       y = "log-posterior predictive density")
```

If we sum over those means we obtain the $\widehat{elpd}$, the expected log predictive density, indicated as `elpd_loo`. $p_\text{loo}$ (`p_loo`) is the sum of the variances and indicates the flexibility of the model fit (also used as the effective number of parameters). `looic`, or LOO-IC, corresponds to $-2 \times ($`elpd_loo`$-$`p_loo`$)$ (for the deviance scale). To illustrate these, we can calculate them manually

```{r}
log_lik(fit_gaus) %>% 
  as_tibble() %>% 
  pivot_longer(cols = everything(),
               names_to = "obs", 
               values_to = "loglik") %>%
  summarise(mean = mean(loglik), var = var(loglik), .by = obs) %>% 
  summarise(elpd_loo = sum(mean), # sum of mean log lik
            p_loo = sum(var), # effective number of parameters
            looic = -2 * (elpd_loo - p_loo)) %>% # Bayesian deviance
  pivot_longer(everything()) %>% 
  mutate(across(value, round, 1))
```

or directly using `loo` (a small difference isn't unexpected).

@@

The Pareto *k* diagnostic indicates that there are some observations that are very unusual under the single distribution Gaussian model.

@@

```{r}
loo_fit <- loo(fit_gaus, fit_lgaus)
loo_fit$diffs %>% 
  as.data.frame() %>% 
  round(1)
```

The model comparison shows that the model `fit_lgaus` (the model with the log-Gaussian likelihood) is approximately `r round(abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2]),0)` standard errors better than `fit_gaus`. This can be calculated by `abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2])` by normalising the difference over its standard error $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, which returns the *z* or *t* value of the difference between models [@sivula2020uncertainty]. A difference of `r round(abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2]),0)` standard errors is substantial. By comparison, a *t*-value of more extreme than $\mid2\mid$ corresponds to a difference that is significant at a $\alpha$-level of 0.05 (i.e. $p<0.05$) [e.g. @baa08book].

Aside, unlike $R^2$, `elpd_loo` and `looic` have no direct interpretation. This is because the size of `elpd_loo` increases as a function of the number of observations. Generally you can think of `elpd_loo` and `looic` as how many possible futures the model can rule out. Important is the difference between models.

In contrast, `loo_R2` provides a Bayesian $R^2$ that measures the variance in the data that the model can capture. This quantity is not ideal for model comparisons because, similar to the frequentist $R^2$, it is subject to overfitting: a model with more parameters can never have a smaller $R^2$, and $R^2$ will innevitabily increase when we add more parameters to a model. A larger $R^2$ means that a model captures the data well but might make predictions that do not generalise out-of-sample [see chapter 11 in @gelman2020regression; also chapter 6 in @mcelreath2016statistical]. The is an example for Bayesian $R^2$ for the single-distribution log-Gaussian model.

@@

The mixture model captures the writing process as a combination of three parameters: the average typing speed of fluent keystroke transitions ($\beta$), the slowdown for hesitant transitions ($\delta$) and the probability of hesitant transitions ($\theta$). In the model below, these are estimated each level of a categorical predictor `location` for the $i^\text{th}$ keystroke interval.

$$
\begin{align}
\text{iki}_{i} \sim & \theta_\text{location[i],participant[i]} \times \text{log}\mathcal{N}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i],participant[i]}) \times \text{log}\mathcal{N}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
		&\text{where}\\
		&\delta \sim \mathcal{N}(0,1)\\
		&\text{constraint: } \delta > 0
\end{align}
$$

Model takes into account two sources of random error: (1) each participant has an individual fluent typing speed that differs across levels of the categorical predictor and is distribution $u_\text{participant} \sim \mathcal{N}(0, \sigma^2_p)$; (2) each participant has in individual hesitation probability $\theta_\text{participant}$ that differs across levels of the categorical predictor.

The parameters of this equation are visualised in Figure \@ref(fig:mixturemodel). The illustrate shows the dissociation of two underlying processes that the model is estimating on the basis of the data. There are two distributions of which the left one in grey is the distribution of fluent keystroke intervals; the right distribution in green is wider and captures longer keystroke intervals (with some overlap between fluent and disfluent transitions).

```{r mixturemodel, echo = F, fig.cap="What's a mixture model for keystroke intervals? The parameters of the mixture model explained visually. Each of the three parameters can determined for each transition location."}

@@

ggplot(data_mm, aes(x)) +
  scale_x_continuous(labels = exp, breaks = log(c(0, 10, 100, 1000, 10000)), limits = log(c(30, 50000))) +
  scale_y_continuous(limits = c(0, .7)) +
  stat_function(geom = "line", fun = plot_mix_comps, colour = "#000000",
                args = list(alpha[1], sigma[1], lam = 1 - theta[1])) +
  stat_function(geom = "line", fun = plot_mix_comps, colour = "#009E73",
                args = list(alpha2[1], sigmap[1], lam = theta[1])) +
  stat_function(geom = "area", fun = plot_mix_comps, fill = "#000000",
                show.legend = F, alpha = .25, 
                args = list(alpha[1], sigma[1], lam = 1 - theta[1])) +
  stat_function(geom = "area", fun = plot_mix_comps, fill = "#009E73",
                show.legend = F, alpha = .25, 
                args = list(alpha2[1], sigmap[1], lam = theta[1])) +
  annotate("label", x=4,y=.6, label="beta*':'~'Typing speed'", size = 4, parse = T) +
  annotate("label", x=7.25,y=.35, label="atop(theta*':'~'Hesitation probability', '\"Pausing\"')", parse = T, size = 4) +
  annotate("label", x=9,y=.15, parse = T, label="delta*':'~'\"Slowdown\"'", size = 4) +
  geom_segment(x = alpha, y = 0.05, xend = alpha2, yend = 0.05, 
               colour = "blue", alpha = .5,
               arrow = arrow(length = unit(0.2, "cm"), 
                             ends = "both", 
                             type = "closed")) +
  geom_segment(x = alpha, y = 0, xend = alpha, yend = Inf, colour = "darkred", alpha = .5) +
  geom_segment(x = alpha2, y = 0, xend = alpha2, yend = Inf, colour = "darkred", alpha = .5) +
  geom_curve(x = 4, y = .57, xend = alpha - .75, yend = 0.1, 
  arrow = arrow(length = unit(0.03, "npc"))) +
  geom_curve(x = 7.5, y = .3, xend = alpha2 + .85, yend = 0.1,
             curvature=-.3,
  arrow = arrow(length = unit(0.03, "npc"))) +
  geom_curve(x = 9, y = .125, xend = 5.75, yend = 0.0255,
             curvature=-.3,
   arrow = arrow(length = unit(0.03, "npc"))) +
  labs(x = "iki [in msecs]", 
       y = "Density",
       caption = "see Roeser et al. (2021)")

@@

Importantly Figure \@ref(fig:mixturemodel) illustrates the average typing speed of fluent keystroke transitions captured by the $\beta$ parameter; the distance between fluent transitions and disfluent transitions captured by $\delta$ (i.e. the distance between the grey and the green distribution); the hesitation probability or the "pausing frequency" which is the height of the green distribution captured by $\theta$. Most keystrokes are, typically, fluent transitions and hence part of the grey distribution. The frequency of the minority of disfluent transitions is captured by the height of the green distribution. As this model assumes a finite mixture of two distributions, the probability of fluent transitions is the inverse $1-\theta$; in other words, a pausing probability of $\theta=.4$ means that the probability of fluent transitions is $.6$. If the pausing probability decreases to $.3$ the proportion of fluent transitions had increased to $.7$ etc.

## Model input values

The data must be transformed into a list to feed them into Stan. The information required from the data is determined in the Stan code and has to do with what we need to estimate the model parameters and how the model is implemented. The exert from the Stan code below shows the information that is expected as input, which data type is expected (e.g. `int` means integer), their names, and what the smallest (i.e. `lower`) and largest possible value (i.e. `upper`) are. The full Stan code can be found at the end of this tutorial.

@@

Participants and conditions are expected to represent indices in the Stan code. In the code below we use `factor` in combination with `as.numeric` to ensure that there are no empty indices for participants and conditions. We create vectors with participant identifiers `ppts` (numbers $1$ through $I$ where $I$ is the number of participants) and a numeric identifier `condition` for each level of the transition location (levels: before-sentence = 1, before-word = 2, within-word = 3).  The returned values are indices for the parameters in the model: For example, for the location identifier, `beta[1]` is the population estimate for non-hesitant typing intervals at before-sentence transition locations, `beta[2]` for before-word transition locations; `theta[1]` and `theta[2]` are the hesitation probability (on the logit scale) for before-sentence and before-word transition locations, respectively.

@@

## Loading the model

The Stan code can be downloaded from [GitHub](https://github.com/jensroes/prowrite-mixture-models/blob/main/tutorial/stan/molg.stan) using `readLines` and saved locally using `writeLines` as "molg.stan" (mixture of log-Gaussians) in the directory "stan".

```{r eval = F}
# Download Stan code
file <- readLines("https://raw.githubusercontent.com/jensroes/prowrite-mixture-models/main/tutorial/stan/molg.stan")
# Save Stan code in "stan" directory
writeLines(file, "stan/molg.stan")
```

The Stan model can be loaded and translated into an S4 class object using the `stan_model` function and assigned to `molg`. The "molg.stan" file needs to be available in the directory "stan" located inside the current working directory.

@@

This code fits the posterior of a categorical predictor with any number of levels. In other words, the posterior that the model returns can be used to calculate simple differences, main effects and interactions of factorial designs. We demonstrate below how to calculate a simple difference. The same logic can be used to calculate main effects and interactions. Also, participant estimates are calculated for both mixture components and the hesitation probability by location. For including more predictors, or removing, e.g., random error terms, or adjusting priors, the Stan code can be changed by opening it in R or a text editor. The Stan code used is largely based on @sorensen2016bayesian and @vasishth2017. @sorensen2016bayesian presents a detailed tutorial on how to write models in Stan [see also @lambert2018student].

The model is fitting a number of parameters of which not all are interesting for inference. To reduce the size of the model object that contains the posterior we can determine which parameters should be saved and disregard all parameters.

@@

The `sampling` function applies the model to the data using the information introduced before (iterations, warmup, chains, cores, parameters `pars` to be saved). `save_warmup` is set to `FALSE` to discard the warmup samples (which are not used for inference anyway) to reduce the size of the model object. To allow reproducibility (and because Bayesian models involves random number generation) we set the seed using an arbitrary number. Lastly, the control argument was specified with higher values for `adapt_delta` and `max_treedepth`: using higher values here means the model runs slower but means more careful parameter estimation. The model is assigned to `fit_molg`.

@@

As above we save the results using `saveRDS`.

@@

Running this model will take a while to complete sampling depending on your hardware specifications. The time it took my laptop to complete this job can me viewed using `get_elapsed_time`. It is worth to not use all cores to run this model or to use a dedicated high performance machine.

```{r}
get_elapsed_time(fit_molg) / 60 # in mins
```

Inferred posterior estimates shown as average and 95% probability interval can be extracted using the `print` function for each transition location and the three parameter values of interest. `prob` is the $\theta$ parameter on the probability scale rather than on the logit scale. $\beta$ and $\delta$ are on the log scale.

@@

We will return to the interpretation of these estimates below.

Before we interpret these result we need to ensure that the model reached convergence. Convergence can be established using two simple techniques. First, we can inspect the chains in trace plots which show the parameter estimate across iterations (here after warmup) for each chain in different colours. We will look at the population-level parameters saved in `pars`.

To create trace plots, we can apply the `stan_trace` function to the model `fit_molg` and extract the MCMC chains for the parameters in `pars`. The `alpha` argument makes the colours slightly more transparent otherwise it is difficult to see the chains hiding in the back. If the chains overlap and look like "fat hairy caterpillars", chains have converged on the same target distribution. The chains below look good except for one divergent transition for `beta[3]`.

Second, we can calculate the $\hat{R}$ statistic. Successful convergence is reflected in $\hat{R}$ values smaller than 1.05 [@gelman1992]. $\hat{R}$ is similar to the *F* statistic in standard analysis of variance: it tells us how much bigger the variability between chains is compared to the variability within all chains. A value of $\approx 1$ indicates that the variability is essentially identical signifying convergence. Larger $\hat{R}$ values indicate that the chains are diverging. The `rhat` function applied to the model and the parameters of interest extracts the $\hat{R}$ statistic. There is no evidence for convergence problems.

Convergence problems can have many reasons and therefore many solutions: for example, running longer chains (in particular longer warmups), increasing the maximum treedepth and the average acceptance probability (`adapt_delta`), specifying starting values, using sensible regulating priors, constraining parameters, or changing the parametrisation of the model. Severe convergence problems indicate model misspecifications.

The model is simulating hypothetical data sets on the basis of the estimated posterior parameter values. This happened for every iteration for every chains. In other words there is a total of `r scales::comma(total_samples)` simulated data sets (and samples per parameter) available to us because

Using the `as.matrix` function we extract a matrix of `y_tilde`, the simulated data. This returns a matrix of the size `total_samples` $\times$ `nrow(data)` (so `r scales::comma(total_samples)` $\times$ `r scales::comma(nrow(data))`). We use the `sample` function to draw `N` randomly sampled hypothetical data sets.

The `ppc_dens_oberlay` function from the `bayesplot` package is then mapping the observed data ($y$; thick blue line) to the simulated data ($y_{rep}$; thin light blue lines). We find that the simulated data sets closely align with the observed data.

For comparison we also visualised the fit for the simulated data of the Gaussian and the log-Gaussian model in Figure \@ref(fig:fitother). Both models show a substantially poorer fit to the keystroke data than the mixture model-based simulations. Note that the Gaussian model does not rule out negative values which are conceptually implausible.

```{r fitother, echo=FALSE, fig.cap="Fit of simulated datasets compared to observed data for the Gaussian and the log-Gaussian model. x-axes were truncated for visability."}
y_tilde <- posterior_predict(fit_gaus, ndraws = N)
p_gaus <- ppc_dens_overlay(fit_gaus$data$iki, y_tilde) +
  labs(subtitle = "Gaussian model") +
  scale_x_continuous(limits = c(-1000, 3000))

y_tilde <- posterior_predict(fit_lgaus, ndraws = N)
p_lgaus <- ppc_dens_overlay(fit_lgaus$data$iki, y_tilde) +
  labs(subtitle = "log-Gaussian model") +
  scale_x_continuous(limits = c(0, 3000))

p_gaus + p_lgaus +
  plot_layout(guides = "collect") 
```

# Model comparsion II

Similar to the section "Model comparison I" we will use leave-one-out cross-validation to compare the three models, the mixture of log-Gaussians, and the single distribution Gaussian and log-Gaussian models. Because the mixture model was fitted directly in Stan and the single-distribution models were fitted in `brms`, we need to extract LOO statistics before for which we need the log-likelihood and the relative effective sample size for the Stan model. We can then go ahead and compare the models using functions provided by the `loo` package .

@@

As described above we used leave-one-out cross-validation to compare the mixture of log-Gaussians model to the single distribution Gaussian and log-Gaussian models. This time we need to use the `loo_compare` function because we had to extract the LOO statistics.

@@

We can then extract the model comparisons and fit statistics:

@@

The difference between the model in the top row and the two other models is shown in `elpd_diff` ($\Delta\widehat{elpd}$) with standard error shown in `se_diff`. Models are ordered from the model with the highest predictive performance to the lowest (i.e. `elpd_loo`). The model comparison shows that the model `fit_molg` (the mixture of log-Gaussian ) is approximately `r round(abs(loo_comp[2,1] / loo_comp[2,2]),0)` standard error better than the single distribution log-Gaussian `fit_lgaus` (try `abs(loo_comp[2,1] / loo_comp[2,2])`). This constitutes substantial support for the mixture model over the single distribution log-Gaussian.

# Posterior probability distribution

At the core of Bayesian inference is the posterior probability distribution. For each model parameter we have `r scales::comma(total_samples)` posterior samples that render the posterior probability distribution. These samples are the result of the posterior sample from the target distribution after warmup across all chains. If the target distribution was not found by the end of the warmup, inference will not be reliable. We know whether the target distribution was determined if the chains converged.

@@

This calculation and visualisation allows us to identify participants that do or do not differ in terms of hesitation frequency depending on transition locations.
